Locality and Effect Reflection

### 2025-09-17 paper.tex
Tighten up some text in Section 2

We can provide effect reflection as a simple library API in OxCaml. In
this section we describe this API and show how to use it to write a
number of typical examples of algebraic effects.

@@

The effect reflection API relies on OxCaml's support for tracking
locality in types, so let's start with an overview of that support.

@@

Locality tracks whether a value can escape the current
\emph{region}. By default, the current region ends when a function
returns, so local values cannot be returned from functions.
\begin{lstlisting}[style=oxcaml]
  let leak_local x =
    let (o @ local) = Some x in
    o
  (*@ \textcolor{red}{Error: o} @*)
         (*@ \textcolor{red}{\textasciicircum} @*)
  (*@ \textcolor{red}{This value escapes its region.} @*)
\end{lstlisting}
The \lstinline[style=oxcaml]{exclave} construct allows functions to
return local values.
\begin{lstlisting}[style=oxcaml]
  let return_local x =
    exclave
      (let (o @ local) = Some x in
       o)
\end{lstlisting}
\lstinline[style=oxcaml]{exclave e} can only appear in tail position
within a function.  It ends the function's region early and then
executes the expression
\lstinline[style=oxcaml]{e}. \lstinline[style=oxcaml]{e} is essentially
executed within the region of the caller of the function and any local
values created within it can be safely returned to that caller.

The place where modes appear within the type algebra is on arrow
types. A function being \lstinline[style=oxcaml]{global} doesn't tell you
that the values it accepts or returns should be
\lstinline[style=oxcaml]{global} too, so instead the arrow type is
labelled with two modes: one for the argument and one for the return
value.
\begin{lstlisting}[style=oxcaml]
  val return_local : 'a @ local -> 'a option @ local
\end{lstlisting}
For convenience, and backwards compatibility with OCaml, either mode can
be omitted, in which case it defaults to
\lstinline[style=oxcaml]{global}.

@@

Note that, whilst there is a \lstinline[style=oxcaml]{global} modality,
there is no \lstinline[style=oxcaml]{local} modality. It does not make
sense for a value which has the ability to leave a region to contain a
value which doesn't, as when the first value leaves a region it will
implicitly cause the second one to leave that region as well. This means
the \lstinline[style=oxcaml]{global} mode is always deep.

@@

The types and operations of our effect reflection API are parameterized
by type constructors representing the operations of an algebraic
effect. In OxCaml, such higher-kinded interfaces must be built using
functors, so we start with a simple module type
\lstinline[style=oxcaml]{Op} to act as the parameter type for these
functors:
\begin{lstlisting}[style=oxcaml]
  module type Op = sig type 'a t end
\end{lstlisting}
The fundamental types of the API are the \lstinline[style=oxcaml]{Term}
and \lstinline[style=oxcaml]{Handler} types.
\begin{lstlisting}[style=oxcaml]
  module Term (O : Op) : sig
    type 'a t =
      | Return : 'a @@ global -> 'a t
      | Op : 'r O.t @@ global * ('r -> 'a t) -> 'a t
  end

@@

For an effect signature \lstinline[style=oxcaml]{O},
\lstinline[style=oxcaml]{'a Term(O).t} is the type of terms in the free
algebra of the corresponding effect -- otherwise known as the freer
monad~\cite{kiselyov2015freer}. \lstinline[style=oxcaml]{Handler(O).t}
is the type of mappings from the operations of
\lstinline[style=oxcaml]{O} into the operations of the current ambient
monad. Alternatively, \lstinline[style=oxcaml]{Handler(O).t} can be
described as representing algebraic effect handlers that are handling
the current computation.

The main piece of the API is a family of isomorphisms:
\begin{equation*}
  \text{\lstinline[style=oxcaml]{'a Term(O).t}} \qquad
  \cong \qquad \text{\lstinline[style=oxcaml]{Handler(O).t @ local -> 'a}}
\end{equation*}
Due to the absence of mode polymorphism in OxCaml, we expose this
isomorphism as three functions: \lstinline[style=oxcaml]{reify} and
\lstinline[style=oxcaml]{reify_local} which implement the left-to-right
direction at the \lstinline[style=oxcaml]{global} and
\lstinline[style=oxcaml]{local} modes respectively; and
\lstinline[style=oxcaml]{perform}, which is equivalent to the
right-to-left direction but makes for a more convenient primitive in
practice. \lstinline[style=oxcaml]{perform} amounts to reflection of
individual operations as opposed to full terms.
\begin{lstlisting}[style=oxcaml]
  module Reflection (O : Op) : sig
    val reify : (Handler(O).t @ local -> 'a) -> 'a Term(O).t

@@

    val perform : 'a O.t -> (Handler(O).t @ local -> 'a)
  end
\end{lstlisting}
Using \lstinline[style=oxcaml]{perform} we can implement reflection of
global terms as:
\begin{lstlisting}[style=oxcaml]
  let rec reflect t h =
    match t with
    | Return x -> x
    | Op(op, k) -> reflect (k (perform op h)) h
\end{lstlisting}
with type:
\begin{lstlisting}[style=oxcaml]
  val reflect : 'a Term(O).t -> (Handler(O).t @ local -> 'a)
\end{lstlisting}
Reflection of local terms can be defined similarly.

>>>>

We provide effect reflection as a simple library API in OxCaml. In this
section we describe this API and show how to use it to write typical
examples of algebraic effects.

@@

The effect reflection API relies on OxCaml's support for tracking
locality in types, which we describe here. For a full treatment see
Lorenzen et al.~\cite{lorenzen2024oxidizing}.

@@

Locality tracks whether a value can escape the current
\emph{region}. By default, the current region ends when a function
returns, so local values cannot be returned from functions.
\begin{lstlisting}[style=oxcaml]
  let leak_local x =
    let (o @ local) = Some x in
    o
  (*@ \textcolor{red}{Error: o} @*)
         (*@ \textcolor{red}{\textasciicircum} @*)
  (*@ \textcolor{red}{This value escapes its region.} @*)
\end{lstlisting}
The \lstinline[style=oxcaml]{exclave} construct allows functions to
return local values.
\begin{lstlisting}[style=oxcaml]
  let return_local x =
    exclave
      (let (o @ local) = Some x in
       o)
\end{lstlisting}
\lstinline[style=oxcaml]{exclave e} can only appear in tail position
within a function.  It ends the function's region early and then
executes the expression \lstinline[style=oxcaml]{e}. Since
\lstinline[style=oxcaml]{e} is executed within the region of the caller
of the function, any local values created within it can be safely
returned to that caller.

Arrow types are labelled with two modes: one for the argument and one
for the return value.
\begin{lstlisting}[style=oxcaml]
  val return_local : 'a @ local -> 'a option @ local
\end{lstlisting}
For convenience, and backwards compatibility with OCaml, either mode can
be omitted, in which case it defaults to
\lstinline[style=oxcaml]{global}.

@@

Note that, whilst there is a \lstinline[style=oxcaml]{global} modality,
there is no \lstinline[style=oxcaml]{local} modality. It does not make
sense for a value which has the ability to leave a region to contain a
value which doesn't have that ability: when the first value leaves a
region it will implicitly cause the second one to leave that region as
well. This means the \lstinline[style=oxcaml]{global} mode is always
deep.

@@

The types and operations of our effect reflection API are parameterized
by type constructors representing the operations of an algebraic
effect. In OxCaml, such higher-kinded interfaces must be built using
functors. We start with a simple module type
\lstinline[style=oxcaml]{Op} to act as the parameter type for these
functors:
\begin{lstlisting}[style=oxcaml]
  module type Op = sig type 'a t end
\end{lstlisting}
The fundamental types of the API are the \lstinline[style=oxcaml]{Term}
and \lstinline[style=oxcaml]{Handler} types.
\begin{lstlisting}[style=oxcaml]
  module Term (O : Op) : sig
    type 'a t =
      | Return : 'a @@ global -> 'a t
      | Op : 'r O.t @@ global * ('r -> 'a t) -> 'a t
  end

@@

For an effect signature \lstinline[style=oxcaml]{O},
\lstinline[style=oxcaml]{'a Term(O).t} is the type of terms in the free
algebra of the corresponding effect -- otherwise known as the freer
monad~\cite{kiselyov2015freer}. \lstinline[style=oxcaml]{Handler(O).t}
represents the algebraic effect handlers that are handling the current
computation.

The main piece of the API is a family of isomorphisms:
\begin{equation*}
  \text{\lstinline[style=oxcaml]{'a Term(O).t}} \qquad
  \cong \qquad \text{\lstinline[style=oxcaml]{Handler(O).t @ local -> 'a}}
\end{equation*}
We expose this isomorphism as three functions:
\begin{lstlisting}[style=oxcaml]
  module Reflection (O : Op) : sig
    val reify : (Handler(O).t @ local -> 'a) -> 'a Term(O).t

@@

    val perform : 'a O.t -> (Handler(O).t @ local -> 'a)
  end
\end{lstlisting}
\lstinline[style=oxcaml]{reify} and
\lstinline[style=oxcaml]{reify_local} implement the left-to-right
direction. We need two separate functions because OxCaml lacks any form
of mode polymorphism.

\lstinline[style=oxcaml]{perform} is equivalent to the right-to-left
direction but makes for a more convenient primitive in
practice. \lstinline[style=oxcaml]{perform} amounts to reflection of
individual operations as opposed to full terms. Using
\lstinline[style=oxcaml]{perform} we can implement reflection of global
terms as:
\begin{lstlisting}[style=oxcaml]
  let rec reflect t h =
    match t with
    | Return x -> x
    | Op(op, k) -> reflect (k (perform op h)) h
\end{lstlisting}
with type:
\begin{lstlisting}[style=oxcaml]
  val reflect : 'a Term(O).t -> (Handler(O).t @ local -> 'a)
\end{lstlisting}
Reflection of local terms can be defined similarly.
